Stochastic blockmodeling of relational event dynamics

### 2012-06-01 docs/experiments.tex
Small edits and TODOs in Experiments/Model sections.

Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
The two panels on the left show the variation in $s_0(t_M,i,j)$, $s_1(t_M,i,j)$ and $s_3(t_M,i,j)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
Similarly, block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
The model has learned that this structure exists in the data and our inferences from the parameter estimates are sensible given the observed statistics.

@@

We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood and recall for held-out data.
Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.  I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.

@@

In order to investigate the role of the number of clusters, $K$, we introduce an upper bound on the number of clusters during the fitting process.  In Table ~\ref{tab:results} we compare the loglikelihood on held-out test data while varying the maximum allowed $K$.  A relational event model (i.e. our model with $K=1$) outperforms the baselines on these data sets.  For the synthetic data example introduced in Section \ref{sec:simulation} the fitted models with $K>1$ have predictive performance comparable to the true model; a standard relational event model does not perform as well because it does not have the flexibility to model the dynamics of each block separately.  For the classroom data set we see only a slight improvement with $K>1$, while with the mobile phone calls the $K=3$ model performs best.
In Table~\ref{tab:recall5} and Table~\ref{tab:recall20} we include the corresponding results for the recall experiment at a cutoff of 5 and 20, respectively.  The results there are similar.

>>>>

Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
The two panels on the left show the variation in $s_0(t_M,i,j,\mathcal{A}_t)$, $s_1(t_M,i,j,\mathcal{A}_t)$ and $s_3(t_M,i,j,\mathcal{A}_t)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
Similarly, block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
The model has learned that this structure exists in the data and our inferences from the parameter estimates are sensible given the observed statistics.

@@

We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood and recall for held-out data.
Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\hat{\lambda}_{i,j}(t_m) = \frac{1}{S}\sum_k \lambda_{i,j}(t_m | \boldsymbol{\beta}^{(k)}, \mathbf{z}^{(k)})$ is averaged using $K$ posterior samples given the training data.
% TODO: We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
% TODO: The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.

@@

In order to investigate the role of the number of clusters, we introduce an upper bound $K^*$ on the number of clusters during the fitting process.  In Table ~\ref{tab:results} we compare the loglikelihood on held-out test data while varying the maximum allowed $K^*$.  A relational event model (i.e. our model with $K=1$) outperforms the baselines on these data sets.  For the synthetic data example introduced in Section \ref{sec:simulation} the fitted models with $K>1$ have predictive performance comparable to the true model; a standard relational event model does not perform as well because it does not have the flexibility to model the dynamics of each block separately.  For the classroom data set we see only a slight improvement with $K>1$, while with the mobile phone calls the $K=3$ model performs best.
In Table~\ref{tab:recall5} and Table~\ref{tab:recall20} we include the corresponding results for the recall experiment at a cutoff of 5 and 20, respectively.  The results there are similar.

### 2012-06-01 docs/model.tex
Small edits and TODOs in Experiments/Model sections.

We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta_{\sigma})$.
The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
% TODO: Expand on this?

@@

Table \ref{tab:stats}  lists the statistics  $\mathbf{s}(t,i,j,\mathcal{A}_t)$ used in Section \ref{sec:experiments}.
For example, a particular node sending often may indicate they will continue to be a sender in the future.
We normalize these counts by the number of events up until a dyad's prior changepoint. %\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.}
Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

@@

The proposed family of models generalizes several important special cases.
For example,  using only the intercept statistic $s_0(t,i,j) = 1$ is analogous to the stochastic block model for static networks.
Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
As these intensities do not change under this specification, the likelihood simplifies to
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
\begin{align}
\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
\label{eqn:multllk}
\end{align}
The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1984}, though here the possible choices are all the dyads in $\mathcal{R}$.

\begin{table}[t]
\footnotesize
\center
\begin{tabular}{|l|l|}
\hline
Statistic & Formula\\
\hline
\hline
Intercept& $s_{0}(t,i,j) = 1$\\
Reciprocity (AB-BA)& $s_{1}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
Turn-continuing (AB-AY)& $s_{2}(t,i,j) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
Turn-taking (AB-BY)&$s_{3}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
Sender out-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender in-degree& $s_{5}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
Dyad count& $s_{6}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
\hline
\end{tabular}
\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{ij}$ previous to event $m$.}
\label{tab:stats}
\end{table}

>>>>

We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta_{\sigma})$.
The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
{\color{red} TODO: Expand on this?}

@@

Table \ref{tab:stats}  lists the statistics  $\mathbf{s}(t,i,j,\mathcal{A}_t)$ used in Section \ref{sec:experiments}.
For example, a particular node sending often may indicate they will continue to be a sender in the future.
We normalize these counts by the number of events up until a dyad's prior changepoint by using $f(x) = \log \frac{x+1}{m + N(N-1)}$. %\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.}
Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.

@@

The proposed family of models generalizes several important special cases.
For example,  using only the intercept statistic $s_0(t,i,j,\mathcal{A}_t) = 1$ is analogous to the stochastic block model for static networks.
Under this model each dyad is a homogeneous Poisson process and all dyad intensities $\lambda_{i,j}$ within block $(z_i,z_j)$ have the same intensity, $\exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.
As these intensities do not change under this specification, the likelihood simplifies to
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$

Alternatively, if one models only the order of the events (ignoring the times at which they occur), the likelihood can be written as
\begin{align}
\mathcal{L}_{\mbox{mult}}(\mathcal{A}|\beta) = \prod_{m=1}^M \frac{\lambda_{i_m,j_m}(t_m | \cdot)}{\sum_{(i,j) \in \mathcal{R}} \lambda_{i,j}(t_m | \cdot)}.
\label{eqn:multllk}
\end{align}
The functional form is similar to conditional logit models used for discrete choice data \cite{McFadden1973}, though here the possible choices are all the dyads in $\mathcal{R}$.

\begin{table}[t]
\footnotesize
\center
\begin{tabular}{|l|l|}
\hline
Statistic & Formula\\
\hline
\hline
Intercept& $s_{0}(t,i,j,\mathcal{A}_t) = 1$\\
Reciprocity (AB-BA)& $s_{1}(t,i,j,\mathcal{A}_t) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}=i)$\\
Turn-continuing (AB-AY)& $s_{2}(t,i,j,\mathcal{A}_t) =  I(i_m=i,j_m=j,i_{v_{mij}}=i,j_{v_{mij}}\ne j)$\\
Turn-taking (AB-BY)&$s_{3}(t,i,j,\mathcal{A}_t) = I(i_m=i,j_m=j,i_{v_{mij}}=j,j_{v_{mij}}\ne i)$\\
%Turn-usurping (AB-XA)& $s_{7}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne j,j_{v_{mij}}=i)$\\
%Turn-usurping (AB-XB)&$s_{8}(t,i,j) = I(i_m=i,j_m=j,i_{v_{mij}} \ne i,j_{v_{mij}}=j)$\\
Sender out-degree& $s_{4}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} I(i_m=i) )$\\
Sender in-degree& $s_{5}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} I(j_m=i) )$\\
Dyad count& $s_{6}(t,i,j,\mathcal{A}_t) = f(\sum_{m:t_m<t} I(i_m=i,j_m=j) )$\\
%Receiver out-degree& $s_{3}(t,i,j) = f(\sum_{m:t_m<t} I(i_m=j))$\\
%Receiver in-degree& $s_{4}(t,i,j) = f(\sum_{m:t_m<t} I(j_m=j))$\\
\hline
\end{tabular}
\caption{Statistics used to specify intensity functions using the previous history $\mathcal{A}_t$, where $v_{mij}$ is the index of the changepoint for $\lambda_{i,j}(t)$ previous to event $m$.}
\label{tab:stats}
\end{table}
